% \iffalse meta-comment
%
% Copyright 2025
% The LaTeX Project and any individual authors listed elsewhere
% in this file.
%
% This file is part of the LaTeX base system.
% -——————————————
%
% It may be distributed and/or modified under the
% conditions of the LaTeX Project Public License, either version 1.3c
% of this license or (at your option) any later version.
% The latest version of this license is in
%    https://www.latex-project.org/lppl.txt
% and version 1.3c or later is part of all distributions of LaTeX
% version 2008 or later.
%
% This file has the LPPL maintenance status "maintained".
%
% The list of all files belonging to the LaTeX base distribution is
% given in the file `manifest.txt'. See also `legal.txt' for additional
% information.
%
% The list of derived (unpacked) files belonging to the distribution
% and covered by LPPL is defined by the unpacking scripts (with
% extension .ins) which are part of the distribution.
%
% \fi
% Filename: ltnews41.tex
%
% This is issue 42 of LaTeX News.

\NeedsTeXFormat{LaTeX2e}[2020-02-02]

\documentclass{ltnews}

%% Maybe needed only for Chris' inadequate system:
\providecommand\Dash {\unskip \textemdash}

%% NOTE: Chris' preferred hyphens!
%% \showhyphens{parameters}
%% \hyphenation{because}

\usepackage[T1]{fontenc}

\usepackage{lmodern,url,hologo}

\usepackage{csquotes}
\usepackage{multicol}
\usepackage{color}

\providecommand\hook[1]{\texttt{#1}}

\providecommand\meta[1]{$\langle$\textrm{\itshape#1}$\rangle$}
\providecommand\option[1]{\texttt{#1}}
\providecommand\env[1]{\texttt{#1}}
\providecommand\Arg[1]{\texttt\{\meta{#1}\texttt\}}


\providecommand\eTeX{\hologo{eTeX}}
\providecommand\XeTeX{\hologo{XeTeX}}
\providecommand\LuaTeX{\hologo{LuaTeX}}
\providecommand\pdfTeX{\hologo{pdfTeX}}
\providecommand\MiKTeX{\hologo{MiKTeX}}
\providecommand\CTAN{\textsc{ctan}}
\providecommand\TL{\TeX\,Live}


\providecommand\githubissue[2][]{\ifhmode\unskip\fi
     \quad\penalty500\strut\nobreak\hfill
     \mbox{\small\slshape(%
       \href{https://github.com/latex3/latex2e/issues/\getfirstgithubissue#2 \relax}%
          	    {github issue#1 #2}%
           )}%
     \par\smallskip}
%% But Chris has to mostly disable \href for his TEXPAD app:
%% \def\href #1#2{#2} % Only For Chris' deficient TeX engine

% simple solution right now (just link to the first issue if there are more)
\def\getfirstgithubissue#1 #2\relax{#1}

\providecommand\sxissue[1]{\ifhmode\unskip
     \else
       % githubissue preceding
       \vskip-\smallskipamount
       \vskip-\parskip
     \fi
     \quad\penalty500\strut\nobreak\hfill
     \mbox{\small\slshape(\url{https://tex.stackexchange.com/#1})}\par}

\providecommand\gnatsissue[2]{\ifhmode\unskip\fi
     \quad\penalty500\strut\nobreak\hfill
     \mbox{\small\slshape(%
       \href{https://www.latex-project.org/cgi-bin/ltxbugs2html?pr=#1\%2F\getfirstgithubissue#2 \relax}%
          	    {gnats issue #1/#2}%
           )}%
     \par}

\let\cls\pkg
\providecommand\env[1]{\texttt{#1}}
\providecommand\acro[1]{\textsc{#1}}

\vbadness=1400  % accept slightly empty columns


\let\finalpagebreak\pagebreak % for TUB (if they use it)
\let\finalvspace\vspace       % for document layout fixes

\makeatletter
% maybe not the greatest design but normally we wouldn't have subsubsections
\renewcommand{\subsubsection}{%
   \@startsection {subsubsection}{2}{0pt}{1.5ex \@plus 1ex \@minus .2ex}%
                  {-1em}{\@subheadingfont\colonize}%
}
\providecommand\colonize[1]{#1:}
\makeatother


% Undo ltnews's \verbatim@font with active < and >
\makeatletter
\def\verbatim@font{\normalsize\ttfamily}
\makeatother


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\providecommand\tubcommand[1]{}
\tubcommand{\input{tubltmac}}

\publicationmonth{November}
\publicationyear{2025  --- DRAFT version for upcoming release}

\publicationissue{42}

\begin{document}

\maketitle
{\hyphenpenalty=10000 \exhyphenpenalty=10000 \spaceskip=3.33pt \hbadness=10000
\tableofcontents}

\setlength\rightskip{0pt plus 3em}

\medskip

\section{Introduction}

\emph{to write}


\section{News from the Tagged PDF project}

\subsection{Normalizing key names for block environments}

\emph{to write}


\section{New or improved commands}

\subsection{Support separate font families for script fonts}

In \TeX's math processing separate fonts can be selected for text, script and
scriptscript sizes. \LaTeX's NFSS traditionally uses the same font family at
different sizes, handling adjustment needed for making fonts appear better in a
script location through the use of optical sizes. This works great for
traditional \TeX{} fonts, but for OpenType fonts this leads to issues. OpenType
MATH assumes the font in a script location has separate features set and
therefore received specific adjustments.

To support this without relying on heuristics based on the font size, a new
command \verb+\DeclareMathScriptfontMapping+ has been added. It takes 3 pairs of
encoding/family arguments to indicate that for the first pair when used as the
math main font the second and the third should be used as the script and
scriptscript font, respectively.
%
\githubissue{1707}

\section{Code improvements}

\subsection{Avoid strange warnings about font substitutions}

A font series value such as \texttt{sbc} contains both the weight
(\texttt{sb}, i.e.\ ``semibold'') and the width (\texttt{c},
i.e.\ ``condensed'') of the font. If you want to reset only one of the two to
``medium'' and keep the other, you can use \verb+\fontseries{m?}+ or
\verb+\fontseries{?m}+: The former switches \texttt{sbc} to \texttt{c}, the
latter switches \texttt{sbc} to \texttt{sb}. However, if the resulting series
did not exist, you got strange warnings, e.g.:
\begin{verbatim}
LaTeX Font Warning:
Font shape `OT1/cmss/c/n' undefined
using `OT1/cmss/m?/n' instead on input line 7.
LaTeX Font Warning:
Font shape `OT1/cmss/m?/n' undefined
using `OT1/cmss/m/n' instead on input line 7.
\end{verbatim}
This has now been corrected so that you get a single, more meaningful warning:
\begin{verbatim}
LaTeX Font Warning:
Font shape `OT1/cmss/c/n' undefined
using `OT1/cmss/m/n' instead on input line 7.
\end{verbatim}
If the \texttt{m} series does not exist either, you will still get strange
warnings, but this should only affect very few fonts. The source file was
also tidied up a little on this occasion.
%
\githubissue{1727}

\subsection{Allow multiple family names in \cs{ProcessKeyOptions}}

The ability to process key--value options was introduced into the kernel in the
June 2022 release~\cite{42:ltnews35}, with the command \cs{ProcessKeyOptions}
carrying out the option assignment. In the original version, this takes an
optional argument which can select one key family (namespace) for options. We
have now extended this to take a comma list of possible families.
%
\githubissue{1756}

\section{Bug fixes}

\subsection{Support active characters correctly with \cs{DeclareRobustCommand}}

The mechanism used by \cs{DeclareRobustCommand} creates an internal command
which has a space added to the name of the document one: so \verb*|\foo | for a
command \verb|\foo|. That fails if applied to an active character: unlike
normal commands, these have to be exactly one character long. Due to the way
the implementation works, to date this would result in redefining \verb*|\ |
every time \cs{DeclareRobustCommand} was used with an active character. This
has now been corrected: robust active characters are now created using the
engine \cs{protected} mechanism and do not use an internal auxiliary. They
still work in file names and labels to give the character itself.
%
\githubissue{345}

%\subsection{A fix}

%% Some text
%
%%\githubissue{XXXX}

%\section{Changes to packages in the \pkg{amsmath} category}

%\section{Changes to packages in the \pkg{graphics} category}

\section{Changes to packages in the \pkg{tools} category}

\subsection{Updating the status of some components}

The \pkg{tools} bundle contains a range of packages with different usage
profiles. Some of these were necessary in the transition from \LaTeX{}~2.09 to
\LaTeXe{}, while others are very widely used in current documents (for example
\pkg{array}). We have therefore marked a small number of packages in
\pkg{tools} as \emph{retained only for historical and stability reasons}, and
where relevant pointed to more up-to-date alternatives; the list is:
\begin{itemize}
  \item \pkg{enumerate}: use \pkg{enumitem} instead
  \item \pkg{rawfonts}: retained as part of \LaTeX{}~2.09 support
  \item \pkg{somedefs}: retained as part of \LaTeX{}~2.09 support
  \item \pkg{theorem}: use \pkg{amsthm} instead
  \item \pkg{verbatim}: use \pkg{fancyvrb} instead
\end{itemize}

%\section{Changes to files in the \pkg{cyrillic} category}

\begin{thebibliography}{9}\frenchspacing

%\fontsize{9.3}{11.3}\selectfont

\bibitem{42:Lamport}
Leslie Lamport.
\newblock \emph{{\LaTeX}: {A} Document Preparation System: User's Guide and Reference
  Manual}.
\newblock \mbox{Addison}-Wesley, Reading, MA, USA, 2nd edition, 1994.
\newblock ISBN 0-201-52983-1.
\newblock Reprinted with corrections in 1996.

\bibitem{42:ltnews} \LaTeX{} Project Team.
  \emph{\LaTeXe{} news 1--42}. November 2025.
  \url{https://latex-project.org/news/latex2e-news/ltnews.pdf}

\bibitem{42:ltnews35} \LaTeX{} Project Team.
  \emph{\LaTeXe{} news 35}. June 2022.
  \url{https://latex-project.org/news/latex2e-news/ltnews35.pdf}

\bibitem{42:ltnews41} \LaTeX{} Project Team.
  \emph{\LaTeXe{} news 41}. June 2025.
  \url{https://latex-project.org/news/latex2e-news/ltnews41.pdf}

\end{thebibliography}

\end{document}
