% \iffalse meta-comment
%
% Copyright (C) 2020-2021
% The LaTeX Project and any individual authors listed elsewhere
% in this file.
%
% This file is part of the LaTeX base system.
% -------------------------------------------
%
% It may be distributed and/or modified under the
% conditions of the LaTeX Project Public License, either version 1.3c
% of this license or (at your option) any later version.
% The latest version of this license is in
%    http://www.latex-project.org/lppl.txt
% and version 1.3c or later is part of all distributions of LaTeX
% version 2008 or later.
%
% This file has the LPPL maintenance status "maintained".
%
% The list of all files belonging to the LaTeX base distribution is
% given in the file `manifest.txt'. See also `legal.txt' for additional
% information.
%
% The list of derived (unpacked) files belonging to the distribution
% and covered by LPPL is defined by the unpacking scripts (with
% extension .ins) which are part of the distribution.
%
% \fi
% Filename: usrguide3.tex

\documentclass{ltxguide}

\usepackage[T1]{fontenc}  % needed for \textbackslash in tt

\title{New \LaTeX\ methods for authors (starting 2020)}
\author{\copyright~Copyright 2020-2021, \LaTeX\ Project Team.\\
   All rights reserved.}

\date{2021-06-11}

\NewDocumentCommand\cs{m}{\texttt{\textbackslash\detokenize{#1}}}
\NewDocumentCommand\marg{m}{\arg{#1}}
\NewDocumentCommand\meta{m}{\ensuremath{\langle}\textit{#1}\ensuremath{\rangle}}
\NewDocumentCommand\pkg{m}{\textsf{#1}}
\NewDocumentCommand\text{m}{\ifmmode\mbox{#1}\else#1\fi}
% Fix a 'feature'
\makeatletter
\renewcommand \verbatim@font {\normalfont \ttfamily}
\makeatother

\begin{document}

\maketitle

\tableofcontents

\section{Introduction}

\LaTeXe{} was released in 1994 and added a number of then-new concepts to
\LaTeX{}. These are described in \texttt{usrguide}, which has largely remained
unchanged. Since then, the \LaTeX{} team have worked on a number of ideas,
firstly a programming language for \LaTeX{} (\pkg{expl3}) and then a range of
tools for document authors which build on that language. Here, we describe
\emph{stable} and \emph{widely-usable} concepts that have resulted from that
work. These `new' ideas have been transferred from development packages
into the \LaTeXe{} kernel. As such, they are now available to \emph{all}
\LaTeX{} users and have the \emph{same stability} as any other part of the
kernel. The fact that `behind the scenes' they are built on \pkg{expl3}
is useful for the development team, but is not directly important to users.

\section{Creating document commands and environments}

\subsection{Overview}

Creating document commands and environments using the \LaTeX3 toolset is based
around the idea that a common set of descriptions can be used to cover almost
all argument types used in real documents. Thus parsing is reduced to a simple
description of which arguments a command takes: this description provides the
`glue' between the document syntax and the implementation of the
command.

First, we will describe the argument types, then move on to explain how these
can be used to create both document commands and environments. Various more
specialized features are then described, which allow an even richer application
of a simple interface set up.

The details here are intended to help users create document commands in
general. More technical detail, suitable for \TeX{} programmers, is included
in \texttt{interface3}.

\subsection{Describing argument types}

In order to allow each argument to be defined independently, the parser does
not simply need to know the number of arguments for a function, but also the
nature of each one. This is done by constructing an \emph{argument
specification}, which defines the number of arguments, the type of each
argument and any additional information needed for the parser to read the user
input and properly pass it through to internal functions.

The basic form of the argument specifier is a list of letters, where each
letter defines a type of argument. As will be described below, some of the
types need additional information, such as default values. The argument types
can be divided into two, those which define arguments that are mandatory
(potentially raising an error if not found) and those which define optional
arguments. The mandatory types
\begin{itemize}
  \item[\texttt{m}] A standard mandatory argument, which can either be a single token
    alone or multiple tokens surrounded by curly braces |{}|. Regardless of the
    input, the argument will be passed to the internal code without the outer
    braces. This is the type specifier for a normal \TeX{} argument.
  \item[\texttt{r}] Given as \texttt{r}\meta{token1}\meta{token2}, this denotes a
    `required' delimited argument, where the delimiters are
    \meta{token1} and \meta{token2}. If the opening delimiter \meta{token1} is
    missing, the default marker |-NoValue-| will be inserted after a suitable
    error.
  \item[\texttt{R}] Given as \texttt{R}\meta{token1}\meta{token2}\marg{default},
    this is a `required' delimited argument as for~\texttt{r},
    but it has a user-definable recovery \meta{default} instead of
    |-NoValue-|.
  \item[\texttt{v}] Reads an argument `verbatim', between the following
    character and its next occurrence, in a way similar to the argument
    of the \LaTeXe{} command \cs{verb}. Thus a \texttt{v}-type argument
    is read between two identical characters, which cannot be any of |%|, |\|,
    |#|, |{|, |}| or \verb*| |. The verbatim argument can also be enclosed
    between braces, |{| and |}|. A command with a verbatim argument will
    produce an error when it appears within an argument of another function.
  \item[\texttt{b}] Only suitable in the argument specification of an
    environment, it denotes the body of the environment, between
    |\begin|\marg{environment} and |\end|\marg{environment}.  See
    Section~\ref{sec:cmd:body} for details.
\end{itemize}
The types which define optional arguments are:
\begin{itemize}
  \item[\texttt{o}] A standard \LaTeX{} optional argument, surrounded with square
    brackets, which will supply the special |-NoValue-| marker if not given
    (as described later).
  \item[\texttt{d}] Given as \texttt{d}\meta{token1}\meta{token2}, an optional
    argument which is delimited by \meta{token1} and \meta{token2}. As with
    \texttt{o}, if no value is given the special marker |-NoValue-| is
    returned.
  \item[\texttt{O}] Given as \texttt{O}\marg{default}, is like \texttt{o}, but returns
    \meta{default} if no value is given.
  \item[\texttt{D}] Given as \texttt{D}\meta{token1}\meta{token2}\marg{default}, it is
    as for \texttt{d}, but returns \meta{default} if no value is given.
    Internally, the \texttt{o}, \texttt{d} and \texttt{O} types are
    short-cuts to an appropriated-constructed \texttt{D} type argument.
  \item[\texttt{s}] An optional star, which will result in a value
    \cs{BooleanTrue} if a star is present and \cs{BooleanFalse}
    otherwise (as described later).
  \item[\texttt{t}] An optional \meta{token}, which will result in a value
    \cs{BooleanTrue} if \meta{token} is present and \cs{BooleanFalse}
    otherwise. Given as \texttt{t}\meta{token}.
  \item[\texttt{e}] Given as \texttt{e}\marg{tokens}, a set of optional
    \emph{embellishments}, each of which requires a \emph{value}.
    If an embellishment is not present, |-NoValue-| is returned.  Each
    embellishment gives one argument, ordered as for the list of
    \meta{tokens} in the argument specification.  All \meta{tokens}
    must be distinct.
  \item[\texttt{E}] As for \texttt{e} but returns one or more \meta{defaults}
    if values are not given: \texttt{E}\marg{tokens}\marg{defaults}. See
    Section~\ref{sec:cmd:embellishment} for more details.
\end{itemize}

\subsection{Modifying argument descriptions}

In addition to the argument \emph{types} discussed above, the argument
description also gives special meaning to three other characters.

First, \texttt{+} is used to make an argument long (to accept paragraph
tokens). In contrast to \cs{newcommand}, this applies on an
argument-by-argument basis. So modifying the example to `|s o o +m O{default}|'
means that the mandatory argument is now \cs{long}, whereas the optional
arguments are not.

Secondly, \texttt{!} is used to control whether spaces are allowed before
optional arguments. There are some subtleties to this, as \TeX{} itself
has some restrictions on where spaces can be `detected': more detail
is given in Section~\ref{sec:cmd:opt-space}.

Finally, the character \texttt{>} is used to declare so-called
`argument processors', which can be used to modify the contents of an
argument before it is passed to the macro definition. The use of argument
processors is a somewhat advanced topic, (or at least a less commonly used
feature) and is covered in Section~\ref{sec:cmd:processors}.

\subsection{Creating document commands and environments}

\begin{decl}
  |\NewDocumentCommand|     \arg{cmd} \arg{arg spec} \arg{code} \\
  |\RenewDocumentCommand|   \arg{cmd} \arg{arg spec} \arg{code} \\
  |\ProvideDocumentCommand| \arg{cmd} \arg{arg spec} \arg{code} \\
  |\DeclareDocumentCommand| \arg{cmd} \arg{arg spec} \arg{code}
\end{decl}
This family of commands are used to create a \meta{cmd}. The argument
specification for the function is given by \meta{arg spec}, and the command
uses the \meta{code} with |#1|, |#2|, etc.\ replaced by the arguments found
by the parser.

An example:
\begin{verbatim}
  \NewDocumentCommand\chapter{s o m}
    {%
      \IfBooleanTF{#1}%
        {\typesetstarchapter{#3}}%
        {\typesetnormalchapter{#2}{#3}}%
    }
\end{verbatim}
would be a way to define a \cs{chapter} command which would essentially behave
like the current \LaTeXe{} command (except that it would accept an optional
argument even when a \texttt{*} was parsed). The \cs{typesetnormalchapter}
could test its first argument for being |-NoValue-| to see if an optional
argument was present. (See Section~\ref{sec:cmd:special} for details of
\cs{IfBooleanTF} and testing for |-NoValue-|.)

The difference between the \cs{New...} \cs{Renew...}, \cs{Provide...}
and \cs{Declare...} versions is the behavior if \meta{cmd} is already
defined.
\begin{itemize}
 \item \cs{NewDocumentCommand} will issue an error if \meta{cmd}
   has already been defined.
 \item \cs{RenewDocumentCommand} will issue an error if \meta{cmd}
   has not previously been defined.
 \item \cs{ProvideDocumentCommand} creates a new definition for
   \meta{function} only if one has not already been given.
 \item \cs{DeclareDocumentCommand} will always create the new
   definition, irrespective of any existing \meta{cmd} with the
   same name.  This should be used sparingly.
\end{itemize}

\begin{decl}
  |\NewDocumentEnvironment|     \arg{env} \arg{arg spec} \arg{beg-code} \arg{end-code} \\
  |\RenewDocumentEnvironment|   \arg{env} \arg{arg spec} \arg{beg-code} \arg{end-code} \\
  |\ProvideDocumentEnvironment| \arg{env} \arg{arg spec} \arg{beg-code} \arg{end-code} \\
  |\DeclareDocumentEnvironment| \arg{env} \arg{arg spec} \arg{beg-code} \arg{end-code}
\end{decl}
These commands work in the same way as \cs{NewDocumentCommand}, etc.\@, but
create environments (\cs{begin}\arg{env} \ldots{}
\cs{end}\arg{env}). Both the \meta{beg-code} and \meta{end-code}
may access the arguments as defined by \meta{arg spec}. The arguments will be
given following \cs{begin}\arg{env}.

\subsection{Optional arguments}
\label{sec:cmd:opt}

In contrast to commands created using \LaTeXe{}'s \cs{newcommand}, optional
arguments created using \cs{NewDocumentCommand} may safely be nested. Thus for
example, following
\begin{verbatim}
\NewDocumentCommand\foo{om}{I grabbed `#1' and `#2'}
\NewDocumentCommand\baz{o}{#1-#1}
\end{verbatim}
using the command as
\begin{verbatim}
\foo[\baz[stuff]]{more stuff}
\end{verbatim}
will print
\begin{quote}
I grabbed `stuff-stuff' and `more stuff'
\end{quote}
This is particularly useful when placing a command with an optional argument
\emph{inside} the optional argument of a second command.

When an optional argument is followed by a mandatory argument with the same
delimiter, the parser issues a warning because the optional argument could not
be omitted by the user, thus becoming in effect mandatory. This can apply to
\texttt{o}, \texttt{d}, \texttt{O}, \texttt{D}, \texttt{s}, \texttt{t},
\texttt{e}, and \texttt{E} type arguments followed by \texttt{r} or
\texttt{R}-type required arguments.

The default for \texttt{O}, \texttt{D} and \texttt{E} arguments can be
the result of grabbing another argument. Thus for example
\begin{verbatim}
  \NewDocumentCommand\foo{O{#2} m}
\end{verbatim}
would use the mandatory argument as the default for the leading optional
one.

\subsection{Spacing and optional arguments}
\label{sec:cmd:opt-space}

\TeX{} will find the first argument after a function name irrespective of any
intervening spaces. This is true for both mandatory and optional arguments. So
|\foo[arg]| and \verb*|\foo [arg]| are equivalent. Spaces are also ignored when
collecting arguments up to the last mandatory argument to be collected (as it
must exist). So after
\begin{verbatim}
  \NewDocumentCommand\foo{m o m}{ ... }
\end{verbatim}
the user input |\foo{arg1}[arg2]{arg3}| and \verb*|\foo{arg1} [arg2] {arg3}|
will both be parsed in the same way.
 
The behavior of optional arguments \emph{after} any mandatory arguments is
selectable. The standard settings will allow spaces here, and thus with
\begin{verbatim}
  \NewDocumentCommand\foobar{m o}{ ... }
\end{verbatim}
both |\foobar{arg1}[arg2]| and \verb*|\foobar{arg1} [arg2]| will find an
optional argument. This can be changed by giving the modified |!| in the
argument specification:
\begin{verbatim}
  \NewDocumentCommand\foobar{m !o}{ ... }
\end{verbatim}
where \verb*|\foobar{arg1} [arg2]| will not find an optional argument.

There is one subtlety here due to the difference in handling by \TeX{} of
`control symbols', where the command name is made up of a single
character, such as `\texttt{\textbackslash\textbackslash}'.
Spaces are not ignored by \TeX{} here,
and thus it is possible to require an optional argument directly follow such a
command. The most common example is the use of \texttt{\textbackslash\textbackslash}
in \pkg{amsmath} environments, which in the terms here would be defined as
\begin{verbatim}
  \NewDocumentCommand\\{!s !o}{ ... }
\end{verbatim}

\subsection{`Embellishments'}
\label{sec:cmd:embellishment}

The \texttt{E}-type argument allows one default value per test token. This is
achieved by giving a list of defaults for each entry in the list, for example:
\begin{verbatim}
  E{^_}{{UP}{DOWN}}
\end{verbatim}
If the list of default values is \emph{shorter} than the list of test tokens,
the special |-NoValue-| marker will be returned (as for the \texttt{e}-type
argument). Thus for example
\begin{verbatim}
  E{^_}{{UP}}
\end{verbatim}
has default \texttt{UP} for the |^| test character, but will return the
|-NoValue-| marker as a default for |_|. This allows mixing of explicit
defaults with testing for missing values.


\subsection{Testing special values}
\label{sec:cmd:special}

Optional arguments make use of dedicated variables to return information about
the nature of the argument received.

\begin{decl}
  |\IfNoValueTF| \arg{arg} \arg{true code} \arg{false code} \\
  |\IfNoValueT|  \arg{arg} \arg{true code} \\
  |\IfNoValueF|  \arg{arg} \arg{false code}
\end{decl}
The \cs{IfNoValue(TF)} tests are used to check if \meta{argument} (|#1|,
|#2|, \emph{etc.}) is the special |-NoValue-| marker. For example
\begin{verbatim}
  \NewDocumentCommand\foo{o m}
    {%
      \IfNoValueTF {#1}%
        {\DoSomethingJustWithMandatoryArgument{#2}}%
        {\DoSomethingWithBothArguments{#1}{#2}}%
    }
\end{verbatim}
will use a different internal function if the optional argument
is given than if it is not present.

Note that three tests are available, depending on which outcome
branches are required: \cs{IfNoValueTF}, \cs{IfNoValueT} and
\cs{IfNoValueF}.

As the \cs{IfNoValue(TF)} tests are expandable, it is possible to
test these values later, for example at the point of typesetting or
in an expansion context.

It is important to note that |-NoValue-| is constructed such that it
will \emph{not} match the simple text input |-NoValue-|, i.e.~that
\begin{verbatim}
  \IfNoValueTF{-NoValue-}
\end{verbatim}
will be logically \texttt{false}.
When two optional arguments follow each other (a syntax we typically
discourage), it can make sense to allow users of the command to
specify only the second argument by providing an empty first
argument.  Rather than testing separately for emptiness and for
|-NoValue-| it is then best to use the argument type~|O| with an
empty default value, and simply test for emptiness using the
\pkg{expl3} conditional \cs{tl_if_blank:nTF} or its \pkg{etoolbox}
analogue \cs{ifblank}.

\begin{decl}
  |\IfValueTF| \arg{arg} \arg{true code} \arg{false code} \\
  |\IfValueT|  \arg{arg} \arg{true code} \\
  |\IfValueF|  \arg{arg} \arg{false code}
\end{decl}
The reverse form of the \cs{IfNoValue(TF)} tests are also available
as \cs{IfValue(TF)}. The context will determine which logical
form makes the most sense for a given code scenario.

\begin{decl}
  |\BooleanFalse|  \\
  |\BooleanTrue|
\end{decl}
The \texttt{true} and \texttt{false} flags set when searching for
an optional character (using \texttt{s} or \texttt{t\meta{char}}) have
names which are accessible outside of code blocks.

\begin{decl}
  |\IfBooleanTF| \arg{arg} \arg{true code} \arg{false code} \\
  |\IfBooleanT|  \arg{arg} \arg{true code} \\
  |\IfBooleanF|  \arg{arg} \arg{false code}
\end{decl}
Used to test if \meta{argument} (|#1|, |#2|, \emph{etc.}) is
\cs{BooleanTrue} or \cs{BooleanFalse}. For example
\begin{verbatim}
  \NewDocumentCommand\foo{sm}
    {%
      \IfBooleanTF {#1}%
        {\DoSomethingWithStar{#2}}%
        {\DoSomethingWithoutStar{#2}}%
    }
\end{verbatim}
checks for a star as the first argument, then chooses the action to
take based on this information.

\subsection{Argument processors}
\label{sec:cmd:processors}

Argument processor are applied to an argument \emph{after} it has been grabbed
by the underlying system but before it is passed to \meta{code}. An argument
processor can therefore be used to regularize input at an early stage, allowing
the internal functions to be completely independent of input form. Processors
are applied to user input and to default values for optional arguments, but
\emph{not} to the special |-NoValue-| marker.

Each argument processor is specified by the syntax \texttt{>}\marg{processor}
in the argument specification. Processors are applied from right to left, so
that
\begin{verbatim}
  >{\ProcessorB} >{\ProcessorA} m
\end{verbatim}
would apply \cs{ProcessorA} followed by \cs{ProcessorB} to the tokens grabbed
by the \texttt{m} argument.

\begin{decl}
  |\SplitArgument| \arg{number} \arg{token(s)}
\end{decl}
This processor splits the argument given at each occurrence of the
\meta{tokens} up to a maximum of \meta{number} tokens (thus
dividing the input into $\text{\meta{number}} + 1$ parts).
An error is given if too many \meta{tokens} are present in the
input. The processed input is placed inside
$\text{\meta{number}} + 1$ sets of braces for further use.
If there are fewer than \arg{number} of \arg{tokens} in the argument
then |-NoValue-| markers are added at the end of the processed
argument.
\begin{verbatim}
  \NewDocumentCommand \foo {>{\SplitArgument{2}{;}} m}
    {\InternalFunctionOfThreeArguments#1}
\end{verbatim}
If only a single character \meta{token} is used for the split, any
category code $13$ (active) character matching the \meta{token} will
be replaced before the split takes place.
Spaces are trimmed at each end of each item parsed.

The \texttt{E} argument type is somewhat special, because with a single
\texttt{E} in the command declaration you may end up with several
arguments in a command (one formal argument per embellishment token).
Therefore, when an argument processor is applied to an \texttt{E}-type
argument, all the arguments pass through that processor before being fed
to the \meta{code}.  For example, this command
\begin{verbatim}
  \NewDocumentCommand \foo { >{\TrimSpaces} e{_^} }
    { [#1](#2) }
\end{verbatim}
applies \cs{TrimSpaces} to both arguments.

\begin{decl}
  |\SplitList| \arg{token(s)}
\end{decl}
This processor splits the argument given at each occurrence of the
\meta{token(s)} where the number of items is not fixed. Each item is
then wrapped in braces within |#1|. The result is that the
processed argument can be further processed using a mapping function
(see below).
\begin{verbatim}
  \NewDocumentCommand \foo {>{\SplitList{;}} m}
    {\MappingFunction#1}
\end{verbatim}
If only a single character \meta{token} is used for the split, any
category code $13$ (active) character matching the \meta{token} will
be replaced before the split takes place.
Spaces are trimmed at each end of each item parsed.

\begin{decl}
  |\ProcessList| \arg{list} \arg{function}
\end{decl}
To support \cs{SplitList}, the function \cs{ProcessList} is available
to apply a \meta{function} to every entry in a \meta{list}. The
\meta{function} should absorb one argument: the list entry. For example
\begin{verbatim}
  \NewDocumentCommand \foo {>{\SplitList{;}} m}
    {\ProcessList{#1}{\SomeDocumentCommand}}
\end{verbatim}

\begin{decl}
  |\ReverseBoolean|
\end{decl}
This processor reverses the logic of \cs{BooleanTrue} and
\cs{BooleanFalse}, so that the example from earlier would become
\begin{verbatim}
  \NewDocumentCommand\foo{>{\ReverseBoolean} s m}
   {%
      \IfBooleanTF#1%
        {\DoSomethingWithoutStar{#2}}%
        {\DoSomethingWithStar{#2}}%
    }
\end{verbatim}

\begin{decl}
  |\TrimSpaces|
\end{decl}
Removes any leading and trailing spaces (tokens with character code~$32$
and category code~$10$) for the ends of the argument. Thus for example
declaring a function
\begin{verbatim}
  \NewDocumentCommand\foo {>{\TrimSpaces} m}
    {\showtokens{#1}}
\end{verbatim}
and using it in a document as
\begin{flushleft}
  \verb=  =\verb*=\foo{ hello world }=
\end{flushleft}
will show `\verb*=hello world=' at the terminal, with the space at each
end removed. \cs{TrimSpaces} will remove multiple spaces from the ends of
the input in cases where these have been included such that the standard
\TeX{} conversion of multiple spaces to a single space does not apply.

\subsection{Body of an environment}
\label{sec:cmd:body}

While environments |\begin|\marg{environment}\ \dots{}\,|\end|\marg{environment}
are typically used in cases where the code implementing the \meta{environment}
does not need to access the contents of the environment (its `body'),
it is sometimes useful to have the body as a standard argument.

This is achieved by ending the argument specification with~\texttt{b}, which is
a dedicated argument type for this situation. For instance
\begin{verbatim}
  \NewDocumentEnvironment{twice} {O{\ttfamily} +b}
    {#2#1#2} {}
\end{verbatim}
\begin{verbatim}
  \begin{twice}[\itshape]
    Hello world!
  \end{twice}
\end{verbatim}
typesets `Hello world!{\itshape Hello world!}'.

The prefix |+| is used to allow multiple paragraphs in the environment's body.
Argument processors can also be applied to \texttt{b}~arguments. By default,
spaces are trimmed at both ends of the body: in the example there would
otherwise be spaces coming from the ends the lines after |[\itshape]| and
|world!|. Putting the prefix |!| before \texttt{b} suppresses space-trimming.

When \texttt{b} is used in the argument specification, the last argument of the environment declaration (e.g., \cs{NewDocumentEnvironment}), which consists of an \meta{end code} to insert at
|\end|\marg{environment}, is redundant since one can simply put that code at
the end of the \meta{start code}. Nevertheless this (empty) \meta{end code}
must be provided.

Environments that use this feature can be nested.

\subsection{Fully-expandable document commands}

Document commands created using \cs{NewDocumentCommand}, etc.\@, are normally
created so that they do not expand unexpectedly. This is done using engine
features, so is more powerful than \LaTeXe{}'s \cs{protect} mechanism. There
are \emph{very rare} occasion when it may be useful to create functions using a
expansion-only grabber. This imposes a number of restrictions on the
nature of the arguments accepted by a function, and the code it implements.
This facility should only be used when \emph{absolutely necessary}.

\begin{decl}
  |\NewExpandableDocumentCommand|     \arg{cmd} \arg{arg spec} \arg{code} \\
  |\RenewExpandableDocumentCommand|   \arg{cmd} \arg{arg spec} \arg{code} \\
  |\ProvideExpandableDocumentCommand| \arg{cmd} \arg{arg spec} \arg{code} \\
  |\DeclareExpandableDocumentCommand| \arg{cmd} \arg{arg spec} \arg{code}
\end{decl}
This family of commands is used to create a document-level \meta{function},
which will grab its arguments in a fully-expandable manner. The
argument specification for the function is given by \meta{arg spec},
and the function will execute \meta{code}. In  general, \meta{code} will
also be fully expandable, although it is possible that this will
not be the case (for example, a function for use in a table might
expand so that \cs{omit} is the first non-expandable non-space token).

Parsing arguments by pure expansion imposes a number of restrictions on
both the type of arguments that can be read and the error checking
available:
\begin{itemize}
  \item The last argument (if any are present) must be one of the
    mandatory types \texttt{m}, \texttt{r} or \texttt{R}.
  \item The `verbatim' argument type \texttt{v} is not available.
  \item Argument processors (using \texttt{>}) are not available.
  \item It is not possible to differentiate between, for example
    |\foo[| and |\foo{[}|: in both cases the \texttt{[} will be
    interpreted as the start of an optional argument. As a
    result, checking for optional arguments is less robust than
    in the standard version.
\end{itemize}

\subsection{Details about argument delimiters}

In normal (non-expandable) commands, the delimited types look for the
initial delimiter by peeking ahead (using \pkg{expl3}'s |\peek_...|
functions) looking for the delimiter token.  The token has to have the
same meaning and `shape' of the token defined as delimiter.
There are three possible cases of delimiters: character tokens, control
sequence tokens, and active character tokens.  For all practical purposes
of this description, active character tokens will behave exactly as
control sequence tokens.

\subsubsection{Character tokens}

A character token is characterized by its character code, and its meaning
is the category code~(|\catcode|).  When a command is defined, the meaning
of the character token is fixed into the definition of the command and
cannot change.  A command will correctly see an argument delimiter if
the open delimiter has the same character and category codes as at the
time of the definition.  For example in:
\begin{verbatim}
  \NewDocumentCommand { \foobar } { D<>{default} } {(#1)}
\end{verbatim}
\begin{verbatim}
  \foobar <hello> \par
  \char_set_catcode_letter:N <
  \foobar <hello>
\end{verbatim}
the output would be:
\begin{verbatim}
  (hello)
  (default)<hello>
\end{verbatim}
as the open-delimiter |<| changed in meaning between the two calls to
|\foobar|, so the second one doesn't see the |<| as a valid delimiter.
Commands assume that if a valid open-delimiter was found, a matching
close-delimiter will also be there.  If it is not (either by being
omitted or by changing in meaning), a low-level \TeX{} error is raised
and the command call is aborted.

\subsubsection{Control sequence tokens}

A control sequence (or control character) token is characterized by is
its name, and its meaning is its definition.
A token cannot have two different meanings at the same time.
When a control sequence is defined as delimiter in a command,
it will be detected as delimiter whenever the control sequence name
is found in the document regardless of its current definition.
For example in:
\begin{verbatim}
  \cs_set:Npn \x { abc }
  \NewDocumentCommand { \foobar } { D\x\y{default} } {(#1)}
  \foobar \x hello\y \par
  \cs_set:Npn \x { def }
  \foobar \x hello\y
\end{verbatim}
the output would be:
\begin{verbatim}
  (hello)
  (hello)
\end{verbatim}
with both calls to the command seeing the delimiter |\x|.

\subsection{Creating new argument processors}

\begin{decl}
  |\ProcessedArgument| 
\end{decl}
Argument processors allow manipulation of a grabbed argument before it is
passed to the underlying code. New processor implementations may be created
as functions which take one trailing argument, and which leave their result in
the \cs{ProcessedArgument} variable. For example, \cs{ReverseBoolean} is
defined as
\begin{verbatim}
\ExplSyntaxOn
  \cs_new_protected:Npn \ReverseBoolean #1
    {
      \bool_if:NTF #1
        { \tl_set:Nn \ProcessedArgument { \c_false_bool } }
        { \tl_set:Nn \ProcessedArgument { \c_true_bool } }
    }
\ExplSyntaxOff
\end{verbatim}
[As an aside: the code is written in \pkg{expl3}, so we don't have to
  worry about spaces creeping into the definition.]

\subsection{Access to the argument specification}

The argument specifications for document commands and environments are
available for examination and use.

\begin{decl}
  |\GetDocumentCommandArgSpec| \arg{function}         \\
  |\GetDocumentEnvironmentArgSpec| \arg{environment}
\end{decl}
These functions transfer the current argument specification for the
requested \meta{function} or \meta{environment} into the token list
variable \cs{ArgumentSpecification}. If the \meta{function} or
\meta{environment} has no known argument specification then an error
is issued. The assignment to \cs{ArgumentSpecification} is local to
the current \TeX{} group.

\begin{decl}
  |\ShowDocumentCommandArgSpec| \arg{function}         \\
  |\ShowDocumentEnvironmentArgSpec| \arg{environment}
\end{decl}
These functions show the current argument specification for the
requested \meta{function} or \meta{environment} at the terminal. If
the \meta{function} or \meta{environment} has no known argument
specification then an error is issued.

\end{document}
