% \iffalse meta-comment
%
% Copyright 1993 1994 1995 1996 1997 1998 1999 2000 2001 2002 2003 2004 2005 2006 2007 2008 2009
% The LaTeX3 Project and any individual authors listed elsewhere
% in this file.
%
% This file is part of the LaTeX base system.
% -------------------------------------------
%
% It may be distributed and/or modified under the
% conditions of the LaTeX Project Public License, either version 1.3c
% of this license or (at your option) any later version.
% The latest version of this license is in
%    http://www.latex-project.org/lppl.txt
% and version 1.3c or later is part of all distributions of LaTeX
% version 2005/12/01 or later.
%
% This file has the LPPL maintenance status "maintained".
%
% The list of all files belonging to the LaTeX base distribution is
% given in the file `manifest.txt'. See also `legal.txt' for additional
% information.
%
% The list of derived (unpacked) files belonging to the distribution
% and covered by LPPL is defined by the unpacking scripts (with
% extension .ins) which are part of the distribution.
%
% \fi
% Filename: ltnews18.tex
%
% This is issue 18 of LaTeX News.

\documentclass{ltnews}

\usepackage[T1]{fontenc}

%%CCC is it Ok to use 'lmodern' fonts these days?
\usepackage{lmodern,url}

\publicationmonth{September}
\publicationyear{2009}

\publicationissue{19}

\begin{document}

\maketitle

\section{New \LaTeX\ release}

This issue of \LaTeX~News marks the first release of a new version of
\LaTeXe\ since the publication of The \LaTeX\ Companion in 2005--2006.

Just in time for \TeX\ Live 2009, this version is a maintenance
release and introduces no new features. A number of small changes have
been made to correct minor bugs in the kernel and improve various
aspects of some of the \texttt{tools} packages.

%%CCC A bit too much hackers jargon in para 1, but I guess that is our audience
%%  these days (who reading this is going to build anything??):
%%  SVN  repository  codebase  build architecture  test suite  migrated
\TeX\ Live
\section{New code repository}

Since the last \LaTeX\ release, the entire codebase has been migrated
to a public SVN repository\footnote{\url
{http://www.latex-project.org/svnroot/latex2e-public/}}
and the entire build architecture re-written. In fact, it has only
been possible for us to consider a new \LaTeX\ release since earlier
this year when the test suite was finally migrated to the new system.
In the process, a bug in \TeX\ Live was discovered, proving that the
tests are still working.

Now that we can easily generate new packaged versions of the \LaTeXe\
distribution, we expect to be able to roll out bug fixes in a much
more timely manner than over the last few years. New versions should
be distributed yearly with \TeX\ Live. Having said this, the
development of the \LaTeXe\ kernel is slowing down as the bugs become
fewer and more subtle. Remember that we cannot change any of the
underlying architecture of the kernel or any design decisions of the
standard classes because we must preserve backwards compatibility with
legacy documents at all costs.

Even new features cannot be added, because any new documents using
them will not compile in systems (such as journal production engines)
that are generally not updated once they've been proven to work as
necessary.

None of this is to say that we consider \LaTeXe\ to be any less
relevant in the publishing world than in years past. A stable system
is a useful one. Furthermore, the third party package system remains a
steady and solid area of development.
%%CCC expansion of last para:
Moreover, the package system continues to provide a flourishing and stable
means
for thye devlopment of a wide range of extensions.

\section{Babel}

%%CCC not sure that it is so different, is it?  Of course, ldf files are
more common
%%  than other 'config type' files.
One area of the \LaTeXe\ codebase that is still receiving feedback to
be incorporated into the main distribution is the babel system for
multilingual typesetting.

[Fill this in with The Plan.]


\section{The future}

%%CCC I do not like the first phrase (purely personnel)!
%%  Surely kernel development finished many many years ago?
While \LaTeXe\ development draws to a close, the \LaTeX3 project is
seeing new life. Our goals here are to provide a transition from the
\LaTeXe\ document processing model to one that has a more flexible
foundation. Work is continuing in the \textsf{expl3} programming
language and the \textsf{xpackages} for document design. Future
announcements
%%CCC
%for
about \LaTeX3 will be available via the \LaTeX\ Project
website and in TUGboat.

\end{document}
---------------------------------------------------------------------------
The Open University is incorporated by Royal Charter (RC 000391), an exempt
charity in England & Wales and a charity registered in Scotland (SC 038302)

_______________________________________________
Latex-team mailing list
Latex-team@latex-project.org
https://lists.dante.de/mailman/listinfo/latex-team
