% \iffalse meta-comment
%
% Copyright 1993 1994 1995 1996 1997 1998 1999 2000 2001 2002 2003 2004 2005 2006 2007 2008 2009
% The LaTeX3 Project and any individual authors listed elsewhere
% in this file. 
% 
% This file is part of the LaTeX base system.
% -------------------------------------------
% 
% It may be distributed and/or modified under the
% conditions of the LaTeX Project Public License, either version 1.3c
% of this license or (at your option) any later version.
% The latest version of this license is in
%    http://www.latex-project.org/lppl.txt
% and version 1.3c or later is part of all distributions of LaTeX 
% version 2005/12/01 or later.
% 
% This file has the LPPL maintenance status "maintained".
% 
% The list of all files belonging to the LaTeX base distribution is
% given in the file `manifest.txt'. See also `legal.txt' for additional
% information.
% 
% The list of derived (unpacked) files belonging to the distribution 
% and covered by LPPL is defined by the unpacking scripts (with 
% extension .ins) which are part of the distribution.
% 
% \fi
% Filename: ltnews18.tex
%
% This is issue 18 of LaTeX News.

\documentclass
    [lw35fonts]    % uncomment this line to get Palatino
     {ltnews}[2004/02/28]

% \usepackage[T1]{fontenc}


\publicationmonth{December}
\publicationyear{2007}
\publicationissue{18}

\begin{document}

\maketitle

% \raisefirstsection

\section{}

\emph{empty shell}


\section{Updates of required packages}

\section{Work on \LaTeX{} fixes}

\section{Current development}

The title of this section is a little misleading as it actually
describes \emph{current} development. In 1998 the \package{expl3}
bundle of packages was put on \ctan\ to demonstrate a possible \LaTeX3
programming environment. These packages have been lying dormant for some
time while the \LaTeX\ Project Team were preoccupied by other things
such as developing the experimental packages \package{xor},
\package{template}, etc., (and also writing that indispensable
and encyclopaedic volume,\\ 
The \LaTeX{} Companion -- 2nd edition).

In October 2004 work on this code base was resumed with the goal of
some day turning it into a kernel for \LaTeX3.  This work can now
also make full use of the widely accepted \eTeX\ extensions. Currently
two areas are central to this work. 

\begin{itemize} 
\item Extending the
  kernel code of \LaTeX3. 
\item Converting the experimental packages
  such as \package{xor}, \package{template} to use the new
  syntax internally. 
\end{itemize} 

Beware! Development of
\package{expl3} is happening so fast that the descriptions above might
be out of date when you read this! If you wish to see what's going on
then go to
\url{http://www.latex-project.org/code.html} where you can download
 fully working code (we hope!).  

\end{document}

